\section*{\acknowledgments}

% Add entry in the table of contents as section
\addcontentsline{toc}{section}{\acknowledgments}

First, I want to extend my gratitude to Professor Inês Lynce for her unwavering support, for encouraging me to pursue a career in research and for being an outstanding role model. Her invaluable advice has helped me not only during the development of this dissertation, but continuously over the past few~years.

I would like to thank Miguel Neves for his patient explanations, for his very thorough (and very colourful!) text revisions, for always knowing a paper about mostly everything, and for offering me the first “queijada da Graciosa” I ever tried.

I want to thank Miguel Ventura for welcoming me at OutSystems and for sharing his knowledge on regular expressions. I hope when he has a problem and he decides to solve it with a regular expression, he is left with \(n < 2\) problems now!

I am very grateful to Ruben Martins for sharing his expertise in program synthesis and for his advice, for always asking the questions to which I do not have the answers (yet!) and for helping me write in-depth result sections.

I would also like to thank Professor Vasco Manquinho, for his remarkable ability to keep the SAT group's severs running, despite our continuous overusing them.

I would like to thank Ricardo and Catarina for facing this challenge with me. We often came across similar troubles and I have to thank them, above all, for the many uncontrollable fits of laughter we shared, which usually fix them all.
%
I also want to thank Daniel, one of my oldest friends, and Pedro, one of the newest, for sharing their experiences, troubles and successes. They strongly inspire my own.

I would like to thank Vik, Katla and Corisco, my feline companions, always willing to help me chase away the bugs.
%
I want to thank my parents and my sister, for encouraging me, for supporting me, and for patiently waiting until I hand in this dissertation before I go back to continually fixing their email clients and smartphone settings.  They are heroes to everyone, but especially to me.

Finally, I want to thank Duarte for always being by my side.

\bigskip
This work was supported by OutSystems, by national funds through FCT, under project UIDB/50021/2020, and project ANI 045917 funded by FEDER and FCT.

\cleardoublepage